Glyph-maps for visualizing climate data and models

### 2011-09-23 paper.tex
finished perception section

A \emph{visual reference grid} \citep{cleveland:1993a} is fundamental to a good data visualization, in order to compare values mapped to position along a line. For glyph-maps, it is not feasible to make a full set of axes for each icon, because these would dominate the picture. Visual reference grids need to be post-attentive, de-emphasized, and serve as posterior look up after the patterns in the data have been read. For glyph-maps, we need to be able to make the following types of comparisons across icons:

@@

Glyph-maps using trend lines enable the reading of trends across
spatial locations directly from the plot. In comparison, to read trend
from facetted colored maps one needs to spot the difference between
the different maps.  Compare the facetted map in Figure
\ref{fig:facetted-map} with the glyph-map of de-seasonalized data in
Figure \ref{fig:nasa-deseas-glyph}. The glyph-map provides a texturing
over the map. The El Ni\~no event is still primary structure, visible
as the region that has a bump in the middle of the time series, the
hot 1997-1998 period. We can also see that the temperatures are
generally more varied over land. There are a few locations with
dramatically increasing temperature in the later years

Another common approach is to compute the slope at each location code
this as color on the map. Cleveland's \citep{cleveland:1993a}
perceptual tests suggests that the reader can perceive the positional
difference of the lines better than the color differences. Accurate
perception of color differences is difficult.

>>>>

A \emph{visual reference grid} \citep{cleveland:1993a} is fundamental
to a good data visualization, in order to compare values mapped to
position along a line. For glyph-maps, it is not feasible to make a
full set of axes for each icon, because these would dominate the
picture. Visual reference grids need to be post-attentive
\citep{healey}, de-emphasized, and serve as posterior look up after
the patterns in the data have been read. For glyph-maps, we need to be
able to make the following types of comparisons across icons:

@@

Glyph-maps using trend lines enable the reading of trends across
spatial locations directly from the plot. In comparison, to read trend
from facetted colored maps one needs to spot the difference between
the different maps.  Compare the facetted map in Figure
\ref{fig:facetted-map} with the glyph-map of de-seasonalized data in
Figure \ref{fig:nasa-deseas-glyph}. The glyph-map provides a texturing
over the map. The El Ni\~no event is still the primary visble
structure, the region that has a bump in the middle of the time
series, correspondingto the hot 1997-1998 period. We can also see
other temporal features: that the temperatures are generally more
varied over land, there are a few locations with dramatically
increasing temperature in the later years typically in locations that
are the high elevation areas, along the Andes, and Sierra Madre
Occiddental in Mexico.

The glyph-map can be considered to be small multiples as named by
\citet{tufte:2001}, in this case, small multiples of time series. By
laying the time series out on the spatial grid, the main task expected
of the viewer is comparing the temporal patterns from one location to
the next. In contrast, in the small multiples used in the facetted
colored map, the viewer is expected to compare the patterns in the
maps from one year to the next, for corresponding months. (See
\citet{carr:1999} for a discussion on perceptual grouping.) That is,
compare and contrast maps with factted colred map, or compare and
contrast time series using the glyph-map. One draw-back of the
glyph-maps, particularly if linear trend is displayed as an icon, is
they may suffer from the Z\"ollner Illusion \citep{Zollner} which is
visible in Figure \ref{fig:gistemp-pred}. On the other hand, comparing
images require that we play a spot the difference game, which is
perceptually difficult \citet{busey}.

An alternative, to both of these approaches, to focus attention on
specific parts of the data calculate the relevant statistic for each
location and display these values as color on a map. For example, to
study the warming or cooling trends, we might calculate the slope of a
linear model at each location and color tiles across the map by the
value of the slope. From this we might see the increasing temperatures
at higher altitudes in the NASA temperature data. However, we would
fail to see the finer points of the time series, that the increase is
not as simple as a linear increase. In addition, Cleveland's
\citep{cleveland:1993a} perceptual tests suggest that the reader can
perceive the positional difference of the lines better than the color
differences. Accurate perception of color differences is difficult.
